\Title

### 2012-12-23 synmut.tex

\begin{abstract}
\noindent

Intrapatient HIV evolution is goverened by selection on the protein level in
the arms race with the immune system (killer T-cells and antibodies).
Synonymous mutations do not have an immunity-related phenotype and are often
assumed to be neutral. In this paper, we show that synonymous changes in
epitope-rich regions are often deleterious but reach frequencies of order one
via genetic hitchhiking.  We analyze time series of viral sequences from the
V1-C5 part of {\it env}, the envelope gene, within individual hosts and observe
that synonymous derived alleles fix in the viral population much less
frequently than expected from neutral models. Instead, synonymous changes tend
to revert on a time scale of two years, suggesting a selection coefficient of the
order of $-1~$\textperthousand. Extensive computer simulations support these
findings quantitatively. We explore possible biological causes and detect a
negative correlation between fixation of an allele and its involvement in
secondary RNA stem-loop structures, indicating that such structures are
functionally relevant for HIV replication {\it in vivo}. This phenonenon is not
observed in other parts of the HIV genome, in which selective sweeps are less
dense and the genetic architecture less constrained.

>>>>

\begin{abstract}
\noindent
Intrapatient HIV evolution is governed by selection on the protein level in the
arms race with the immune system (killer T-cells and antibodies). Synonymous
mutations do not have an immunity-related phenotype and are often assumed to be
neutral. In this paper, we show that synonymous changes in epitope-rich regions
are often deleterious but still reach high frequencies.  We analyze time
series of viral sequences from the \shankaregion~part of {\it env} within individual
hosts and observe that synonymous derived alleles rarely fix in the
viral population. Simulations suggest that such synonymous mutations
have a (Malthusian) selection coefficient of the order of $-0.001$, and that
they are brought up to high frequency by hitchhiking with neighboring beneficial
nonsynonymous alleles. We detect a negative correlation between fixation of an allele and
its involvement in RNA stem-loop structures.
Deleterious synonymous mutations are observed in other parts of the HIV genome, in which
selective sweeps are less dense and hitch-hiking not as strong.
\end{abstract}
